Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-08-12 Manuscript/MPESubmission/Revision/response.tex

\begin{enumerate}[(1)]
\item The Total Evidence method places living and fossil taxa on the same phylogeny.
\item Here we test the effect of missing morphological data on Total Evidence topologies.
\item The number of living taxa with morphological data has greatest effects on topology.
\item Bayesian inference outperforms Maximum Likelihood for correct clade recovery.
\item We recommend increased sampling effort for morphological data of living taxa.
\end{enumerate}

% -----------------
% Comment 4
% -----------------
\item{\textcolor{blue}{\textbf{Introduction:} Needs stronger emphasis on what specifically is unique about this study compared to what the many other studies on missing data have done and found.
Clearly outlining the questions in this study that are left open from previous research is important because some of the findings in this study suggest that there may be more error associated with missing data than previous research has suggested.}}

We have increased our emphasis on novelty throughout the introduction by more clearly stating the difference between the present study and the literature and more clearly outlining the question of this study, as follows:

\textit{Until now, few attempts have been made to study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework (i.e. using both molecular and morphological data; e.g. Wiens et al., 2005; Manos et al., 2007; Pattinson et al., 2014).
These studies assessed the effect of missing data on topology by either (1) comparing a dataset with missing data to subsets without missing data (Wiens et al. 2005); or (2) removing both molecular and some morphological data from living taxa to create artificial fossils (Manos et al., 2007; Pattinson et al., 2014).
Both approaches have shown that missing data are not a major problem and should not be an obstacle to combining both living and fossil species in the same phylogenies.
The way these studies were conducted, however, means that their conclusions are not generally applicable across all scenarios involving missing data in Total Evidence phylogenies.
For example, using an empirical (rather than simulation based) approach limits their conclusions to studies with similar distributions of data across species in the phylogeny.
Additionally, Wiens et al. (2005) did not include fossil taxa in their analyses, so their results cannot be used to make conclusions about how missing data may influence the placement of fossils. 
Manos et al. (2007) and Pattinson et al. (2014) include fossil taxa, but used the patchiness of the fossil record to determine how to remove data from their matrices. 
Data for living species is unlikely to be missing in this patchy way, instead full molecular data with the complete absence of morphological data are likely (Guillerme and Cooper 2015).
Finally, Manos et al. (2007) and Pattinson et al. (2014), mainly focused on how missing data in fossil taxa affects the placement of fossils, ignoring the effects of missing data in living species.}

@@

% -----------------
% Specific Comment 13
% -----------------
\item{\textcolor{blue}{\textbf{lines 80-103:} (1) the authors need to clearly justify the importance of this study compared the many studies that have investigated the effects of missing data on topology inference.
The justification currently given is not particularly strong and the specific unique questions asked in this study that are currently unanswered should be made clear.
The current argument does not always seem justified by the literature review.
For example, line 81 the authors argue that papers by Wiens and Pattinson examined morphological or molecular matrices separately, but in Pattinson, for example, combined datasets for living species were used to simulate extinct species by deleting the molecular and parts of the morphological datasets from living species.
The authors could argue that a limitation of the Pattinson study, which they address here, was the limited scheme for generating missing data from the datasets.
In this study, the authors perturbed the datasets more thoroughly in a clearly structured experimental design to investigate the effects of eliminating living versus fossil data or all data overall.
That is one aspect of this study that I perceived as new and unique.
(2) Another aspect of this study design that could be new but is not explored here is the effect of the missing data on the morphological substitution rate parameter estimates.
Can you recover the true values from the perturbed datasets?
This study also used a unique way of evaluating the tree inference under missing data scenarios.
These and other aspects of the study uniqueness must be clearer.}}

(1) - We thank the reviewer for this comment and now emphasize both the novel aspects of our study and the questions we are actually testing (see reviewer 1 comment 4 above).

(2) - the second point underlined by the reviewer in this specific comment strikes use as really interesting.
We ran this additional analysis by comparing the estimated $\alpha$ parameter for the morphological rates distribution to the ``true'' $\alpha$ parameter.
We found results that the $\alpha$ parameter is usually slightly underestimated for the various ``missing-data'' trees unless the number of living species with no morphological data are high ($M_L$ = 75\%), then, the $\alpha$ are overestimated.
These results are somewhat in agreement with our topological differences results: the number of living species seems to be a driving missing data parameter for inferring Total Evidence trees.
Nonetheless, we considered this study to already be long and complex.
We therefore think that adding an extra question to this study (i.e. what is the effect of missing data on the morphological substitution rate parameter estimates?) might lose the reader and make an already long paper even longer.
Therefore, we added the results of these analyses to \textbf{Appendix-A Morphological rates estimations results}.
We link to this analysis in the methods as follows:

@@

% -----------------
% Specific Comment 58
% -----------------
\item{\textcolor{blue}{\textbf{Table 1:} The wording of the caption is off - the second paragraph is awkward. 
Are these values the probabilities of overlap between the distributions of the 'best' tree versus trees from each inference method pooled across all combinations of missing data? That is what is described in methods line 397 but it doesn't say so in the caption.
Two things:\\
1)  Comparing to the tables in Appendix C, the distribution of probabilities of overlap between max. likelihood and Bayesian trees are higher in most missing data schemes in tables in the Appendices than the overall values given in Table 1. How could the max. likelihood and Bayesian trees be so different?\\
2)  Pooling across all missing data schemes seems to wash out the effect of tree-inference method. I am not sure what a better way to present these would be, but perhaps some kind of matrix where the cells are the joint probabilities of overlap of each method with each missing data category, and then the marginal probabilities would be the probabilities of overlap for each method, averaged over missing data categories, as well as the marginal probabilities of overlap for each missing data category, averaged across methods.  e.g, RF distance:\\
Missing data:   0   25  50      75      Marginal \\
BI              1   0.9 0.8     0.75    0.863 \\
ML              1   0.9 0.75    0.25    0.725 \\
Marginal        1   0.9 0.78    0.5     \\
This may not be feasible in the text due to space limitations, maybe tables like these could be added to an appendix.}}

We dealt with the caption of the Table 1 in reviewer 1 comment 14 above.

For the other two points:
1) We are not entirely sure if the reviewer means (i) why the two methods display such differences ; or (ii) why, within each method, the combination of parameters displays a difference between each single parameters ($M_LM_FM_C$ vs. $M_L$, $M_F$ and $M_C$).
In the first case, we think that the difference between the methods is due to the fact that the Maximum Likelihood trees are more likely to display ``false positive'' topologies (i.e. clades that are incorrect but can be well supported; or not) as the Bayesian consensus tree, when many data are missing, are more likely to collapse nodes, resulting in trees with less resolution but more ``conserved clades'' (i.e. the clade ((a;b);c)) could be collapsed in (a;b;c) which is still a corrected representation of the relations within the clades rather than ((b;c);a)).

### 2015-08-12 Manuscript/MPESubmission/Revision/TEM_revised_manuscript.tex
Checked manuscript

Although missing data does not appear be a major problem in molecular and morphological matrices separately \citep[\hl{as long as enough data overlaps in each case, and missing data are not phylogeneticaly biased;}][]{wiensmissing2003,Wiens01102005,wiensmissing2006,wiensmissing2008,lemmonthe2009,Sanderson22072011,rouresite-specific2011,pattinsonphylogeny2014}, it may become more of an issue in Total Evidence matrices containing both molecular and morphological data for living and fossil taxa.
This may be particularly problematic as fossil taxa (generally) do not have molecular data, resulting in a large section of missing data in Total Evidence matrices.
\hl{Until now, few attempts have been made to study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework }\citep[\hl{i.e. using both molecular and morphological data;}][]{Wiens01102005,manosphylogeny2007,pattinsonphylogeny2014}.
\hl{These studies assessed the effect of missing data on topology by either (1) comparing a dataset with missing data to subsets without missing data }\citep{Wiens01102005}\hl{; or (2) removing both molecular and some morphological data from living taxa to create artificial fossils }\citep{manosphylogeny2007,pattinsonphylogeny2014}.
\hl{Both approaches have shown that missing data are not a major problem and should not be an obstacle to combining both living and fossil species in the same phylogenies.
The way these studies were conducted, however, means that their conclusions are not generally applicable across all scenarios involving missing data in Total Evidence phylogenies.
For example, using an empirical (rather than simulation based) approach limits their conclusions to studies with similar distributions of data across species in the phylogeny.
Additionally, }\cite{Wiens01102005}\hl{ did not include fossil taxa in their analyses, so their results cannot be used to make conclusions about how missing data may influence the placement of fossils.}
\cite{manosphylogeny2007,pattinsonphylogeny2014}\hl{ include fossil taxa, but used the patchiness of the fossil record to determine how to remove data from their matrices. 
Data for living species is unlikely to be missing in this patchy way, instead full molecular data with the complete absence of morphological data is likely }\citep{GuillermeCooperMissing}.
\hl{Finally, }\cite{manosphylogeny2007,pattinsonphylogeny2014}\hl{ mainly focused on how missing data in fossil taxa affects the placement of fossils, ignoring the effects of missing data in living species.}

@@

We focus on the effect of missing data on our ability to recover a ``best'' tree topology because it is a crucial aspect of a phylogeny in many macroevolutionary studies, for example when trying to elucidate the evolutionary relationships among species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}, or for studying evolutionary transitions \citep[e.g.][]{friedmanexplosive2010}.
Although branch length estimation is also important \citep[namely for timing extinction and/or speciation events; e.g.][]{ronquista2012}, we do not consider branch lengths in this study.
This is partially due to difficulties with simulating branch lengths and topology simultaneously, \hl{but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but using topological constraints }\citep{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}.
Thus understanding the sensitivity of topology to missing \hl{data} are important for assessing the accuracy of tree estimation in the Total Evidence framework. To our knowledge, this question has never been formally assessed.

@@

\hl{We simulated the morphological matrix using the rTraitDisc function from the R package ape v. 3.0-11} \citep{paradisape:2004}\hl{ to generate a matrix of 100 character sites for 51 taxa.}
\hl{We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters (based on an empirical review of published matrices, see Appendix A and Fig. A1 within).}
We then ran an independent discrete character simulation for each character using the ``true'' tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to another \citep{Pagel22011994}.
This method allows us to have only two parameters for each character: the number of states and the evolutionary rate.
For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5.
\hl{We used low evolutionary rate parameters to be consistent with the molecular rate parameters, to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal }\citep{wrightbayesian2014}.
\hl{Topological error has been shown to be minimal at a morphological rate of 0.5 when using the M\textit{kv} model }\citep{lewisa2001,wrightbayesian2014}.
hl{Note, however, that } \cite{wrightbayesian2014}\hl{ have shown that low morphological rates ($<$ 0.5) increase variance in topological error, but we discarded simulations with such topological error by selecting only matrices with a ``fair'' phylogenetic signal }(\citep[\hl{see Estimating phylogenies section below;}][]{zanderminimal2004}\hl{ so this should not influence our results.}

@@

\subsection{Removing data}
\hl{To explore the effect of missing morphological data on topological recovery, we removed various amounts of the ``complete'' matrix to obtain matrices with missing morphological data.
Hereafter, we call these matrices with missing morphological data the ``missing-data'' matrices.
Note that the amount of molecular data remained constant throughout our simulations: 1000 molecular characters for living taxa and no molecular data for fossil taxa (see above).
We removed morphological data using three data incompleteness parameters:}
\begin{enumerate}
\item{\hl{The proportion of missing living taxa ($M_L$).}}
\hl{This first missing-data parameter corresponds to the proportion of living taxa with no morphological data. It represents the number of living taxa that are present in the matrix but have only molecular data available. This reflects the fact that, because of the increasing ease of collecting molecular data, morphological data for living species is rarely collected }\citep{GuillermeCooperMissing}.
\hl{Therefore, many living species will have only molecular data available.
In practice, we removed all the morphological data from randomly chosen living taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of living taxa with no morphological data.}
\item{\hl{The proportion of missing data in the fossil record ($M_F$).}}
\hl{TThis missing data parameter represents the completeness of the fossil record.
Due to preservation biases, missing data for fossil taxa are common }\citep{sansomfossilization2013}.
\hl{In practice, we randomly removed a proportion of data among the fossil taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of overall missing data for the fossil taxa.
Note that 50\% missing data for fossil taxa does not mean that each fossil is missing 50\% of its morphological data.
Instead this 50\% refers to missing fossil data across the whole matrix.
Some fossils may retain 100\% of their data and others may lose all their data at this parameter value (down to a minimum threshold of 5\% available data; see below).}
\item{\hl{The number of morphological characters for both living and fossil taxa ($N_C$).}}
\hl{This parameter is not a missing data parameter \textit{per se} but rather an indication of the size of the matrix.
Any morphological matrix of any size has indeterminate missing data, given that the total number of characters is undefined, but presumably large.
Therefore, this parameter corresponds to the overall number of characters available for both living and fossil taxa.
In practice, we randomly removed whole characters from the morphological matrix reducing it to: 100, 90,75, 50 or 25 characters.
Note that these levels are equivalent to the two other parameters (i.e. 0\%, 10\%, 25\%, 50\% or 75\% of ``missing'' morphological characters).}
\end{enumerate}

\hl{Each parameter represents a different way of removing data from the morphological part of the matrix: $M_L$ removes entire rows from the living data; $M_F$ removes cells from the fossil data; and $N_C$ removes columns across both living and fossil data.
Note that $M_L$ and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa.

We created matrices using all parameter combinations resulting in 125 ($5^3$) ``missing-data'' matrices.
Note that one of these combinations ($M_L$=0\%; $M_F$=0\% and $N_C$=100) has no missing data so is equivalent to the ``complete'' matrix, thus we have one effectively complete matrix in our 125 ``missing-data'' matrices.
In practice, we first removed the data following the two missing data parameters $M_L$ and $M_F$ and then removed data following the $N_C$ parameters.
To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxon.
Note that the living taxa always had at least 90\% of data (the 1000 molecular characters).}

@@

\subsubsection{Robinson-Foulds distance}
The Robinson-Foulds distance \citep{RF1981}, or ``path difference", measures the difference between the number of clades and \hl{twice the number} of shared clades across two trees.
The metric reflects the distance between the distributions of tips among clades in the two trees \citep{RF1981} (see Appendix B for calculation details).
This metric is bounded between zero, when the two trees are identical, and $2(n-2)$ (for two trees with $n$ taxa) when there is no shared clade in the two trees.
This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity.
We normalised this metric following Bogdanowicz's Normalised Tree Similarity (NTS) method \citep{Bogdanowicz2012}.
This method scales any tree comparison metric using the mean distance between 1000 random trees (see Appendix B for the calculation details).
This method is a generalisation of the topological accuracy method \citep{Price2010} allowing to compare topological differences between any tree with any tree comparison metric.
In practice when the Normalised Robinson-Foulds metric between two trees is equal to one, the trees are identical; if the metric is equal to zero, the trees are no more different than expected by chance; finally if the metric is less than zero, the trees are more different than expected by chance.
Note that once rescaled, the Normalised Robinson-Foulds metric is a measure of similarity, rather than of distance like the original Robinson-Foulds metric.

>>>>

Although missing data does not appear be a major problem in molecular and morphological matrices separately \citep[\hl{as long as enough data overlaps in each case, and missing data are not phylogeneticaly biased;}][]{wiensmissing2003,Wiens01102005,wiensmissing2006,wiensmissing2008,lemmonthe2009,Sanderson22072011,rouresite-specific2011,pattinsonphylogeny2014}, it may become more of an issue in Total Evidence matrices containing both molecular and morphological data for living and fossil taxa.
This may be particularly problematic as fossil taxa (generally) do not have molecular data, resulting in a large section of missing data in Total Evidence matrices.
\hl{Until now, few attempts have been made to study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework }\citep[\hl{i.e. using both molecular and morphological data;}][]{Wiens01102005,manosphylogeny2007,pattinsonphylogeny2014}.
\hl{These studies assessed the effect of missing data on topology by either (1) comparing a dataset with missing data to subsets without missing data }\citep{Wiens01102005}\hl{; or (2) removing both molecular and some morphological data from living taxa to create artificial fossils }\citep{manosphylogeny2007,pattinsonphylogeny2014}.
\hl{Both approaches have shown that missing data are not a major problem and should not be an obstacle to combining both living and fossil species in the same phylogenies.
The way these studies were conducted, however, means that their conclusions are not generally applicable across all scenarios involving missing data in Total Evidence phylogenies.
For example, using an empirical (rather than simulation based) approach limits their conclusions to studies with similar distributions of data across species in the phylogeny.
Additionally, }\cite{Wiens01102005}\hl{ did not include fossil taxa in their analyses, so their results cannot be used to make conclusions about how missing data may influence the placement of fossils.}
\cite{manosphylogeny2007,pattinsonphylogeny2014}\hl{ include fossil taxa, but used the patchiness of the fossil record to determine how to remove data from their matrices. 
Data for living species is unlikely to be missing in this patchy way, instead full molecular data with the complete absence of morphological data are likely }\citep{GuillermeCooperMissing}.
\hl{Finally, }\cite{manosphylogeny2007}\hl{ and }\cite{pattinsonphylogeny2014}\hl{ mainly focused on how missing data in fossil taxa affects the placement of fossils, ignoring the effects of missing data in living species.}

@@

We focus on the effect of missing data on our ability to recover a ``best'' tree topology because it is a crucial aspect of a phylogeny in many macroevolutionary studies, for example when trying to elucidate the evolutionary relationships among species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}, or for studying evolutionary transitions \citep[e.g.][]{friedmanexplosive2010}.
Although branch length estimation is also important \citep[namely for timing extinction and/or speciation events; e.g.][]{ronquista2012}, we do not consider branch lengths in this study.
This is partially due to difficulties with simulating branch lengths and topology simultaneously, \hl{but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but using topological constraints }\citep{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}.
Thus understanding the sensitivity of topology to missing data are important for assessing the accuracy of tree estimation in the Total Evidence framework. To our knowledge, this question has never been formally assessed.

@@

\hl{We simulated the morphological matrix using the rTraitDisc function from the R package ape v. 3.0-11} \citep{paradisape:2004}\hl{ to generate a matrix of 100 character sites for 51 taxa.}
\hl{We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters (based on an empirical review of published matrices, see Appendix A and Fig. A1 within).}
We then ran an independent discrete character simulation for each character using the ``true'' tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to another \citep{Pagel22011994}.
This method allows us to have only two parameters for each character: the number of states and the evolutionary rate.
For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5.
\hl{We used low evolutionary rate parameters to be consistent with the molecular rate parameters, to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal }\citep{wrightbayesian2014}.
\hl{Topological error has been shown to be minimal at a morphological rate of 0.5 when using the M\textit{kv} model }\citep{lewisa2001,wrightbayesian2014}.
hl{Note, however, that } \cite{wrightbayesian2014}\hl{ have shown that low morphological rates ($<$ 0.5) increase variance in topological error, but we discarded simulations with such topological error by selecting only matrices with a ``fair'' phylogenetic signal }\citep[\hl{see Estimating phylogenies section below;}][]{zanderminimal2004}\hl{ so this should not influence our results.}

@@

\subsection{Removing data}
\hl{To explore the effect of missing morphological data on topological recovery, we removed various amounts of the ``complete'' matrix to obtain matrices with missing morphological data.
Hereafter, we call these matrices with missing morphological data the ``missing-data'' matrices.
Note that the amount of molecular data remained constant throughout our simulations: 1000 molecular characters for living taxa and no molecular data for fossil taxa (see above).
We removed morphological data using three data incompleteness parameters:}
\begin{enumerate}
\item{\hl{The proportion of missing living taxa ($M_L$).}}
\hl{This first missing-data parameter corresponds to the proportion of living taxa with no morphological data. It represents the number of living taxa that are present in the matrix but have only molecular data available. This reflects the fact that, because of the increasing ease of collecting molecular data, morphological data for living species is rarely collected }\citep{GuillermeCooperMissing}.
\hl{Therefore, many living species will have only molecular data available.
In practice, we removed all the morphological data from randomly chosen living taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of living taxa with no morphological data.}
\item{\hl{The proportion of missing data in the fossil record ($M_F$).}}
\hl{This missing data parameter represents the completeness of the fossil record.
Due to preservation biases, missing data for fossil taxa are common }\citep{sansomfossilization2013}.
\hl{In practice, we randomly removed a proportion of data among the fossil taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of overall missing data for the fossil taxa.
Note that 50\% missing data for fossil taxa does not mean that each fossil is missing 50\% of its morphological data.
Instead this 50\% refers to missing fossil data across the whole matrix.
Some fossils may retain 100\% of their data and others may lose all their data at this parameter value (down to a minimum threshold of 5\% available data; see below).}
\item{\hl{The number of morphological characters for both living and fossil taxa ($N_C$).}}
\hl{This parameter is not a missing data parameter \textit{per se} but rather an indication of the size of the matrix.
Any morphological matrix of any size has indeterminate missing data, given that the total number of characters is undefined, but presumably large.
Therefore, this parameter corresponds to the overall number of characters available for both living and fossil taxa.
In practice, we randomly removed whole characters from the morphological matrix reducing it to: 100, 90,75, 50 or 25 characters.
Note that these levels are equivalent to the two other parameters (i.e. 0\%, 10\%, 25\%, 50\% or 75\% of ``missing'' morphological characters).}
\end{enumerate}

\hl{Each parameter represents a different way of removing data from the morphological part of the matrix: $M_L$ removes entire rows from the living data; $M_F$ removes cells from the fossil data; and $N_C$ removes columns across both living and fossil data.
Note that $M_L$ and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa.}

\hl{We created matrices using all parameter combinations resulting in 125 ($5^3$) ``missing-data'' matrices.
Note that one of these combinations ($M_L$=0\%; $M_F$=0\% and $N_C$=100) has no missing data so is equivalent to the ``complete'' matrix, thus we have one effectively complete matrix in our 125 ``missing-data'' matrices.
In practice, we first removed the data following the two missing data parameters $M_L$ and $M_F$ and then removed data following the $N_C$ parameters.
To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxon.
Note that the living taxa always had at least 90\% of data (the 1000 molecular characters).}

@@

\subsubsection{Robinson-Foulds distance}
The Robinson-Foulds distance \citep{RF1981}, or ``path difference", measures the difference between the number of clades and \hl{twice the number} of shared clades across two trees.
The metric reflects the distance between the distributions of tips among clades in the two trees \citep{RF1981} (see Appendix B for calculation details).
This metric is bounded between zero, when the two trees are identical, and $2(n-2)$ (for two trees with $n$ taxa) when there is no shared clade in the two trees.
This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa \textit{(((a,b),c),((d,e),f))}, the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity.
We normalised this metric following Bogdanowicz's Normalised Tree Similarity (NTS) method \citep{Bogdanowicz2012}.
This method scales any tree comparison metric using the mean distance between 1000 random trees (see Appendix B for the calculation details).
This method is a generalisation of the topological accuracy method \citep{Price2010} allowing to compare topological differences between any tree with any tree comparison metric.
In practice when the Normalised Robinson-Foulds metric between two trees is equal to one, the trees are identical; if the metric is equal to zero, the trees are no more different than expected by chance; finally if the metric is less than zero, the trees are more different than expected by chance.
Note that once rescaled, the Normalised Robinson-Foulds metric is a measure of similarity, rather than of distance like the original Robinson-Foulds metric.
